Preventing Runtime Errors of Redis at Compile Time

### 2015-11-20 paper.tex
codename for the purposes of double-blind reviewing

The database system and the package we are targeting are \emph{Redis} and
 \emph{Hedis} respectively, and our implementation is available as \emph{Edis}
 on \emph{Hackage}.

@@

\begin{itemize}[noitemsep]
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
 language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and constraints of
 an action in types, with indexed monad\cite{indexedmonad} and other language
 extensions such as closed type-families\cite{closedtypefamilies} and
 constraints kinds\cite{constraintskinds}.
\item Edis, a package we built for programmers. This package helps programmers
 to write more reliable Redis programs, and also makes Redis polymorphic by
 automatically converting back and forth from values of arbitrary types and
 boring ByteStrings.
\end{itemize}

@@

We define a new datatype \mintinline{haskell}{Edis}, which is basically just
 the context \mintinline{haskell}{Redis} indexed with more information in types.
 We make it an instance of \mintinline{haskell}{IMonad}.

\begin{minted}[xleftmargin=1em]{haskell}
newtype Edis p q a = Edis { unEdis :: Redis a }

instance IMonad Edis where
    unit = Edis . return
    bind m f = Edis (unEdis m >>= unEdis . f )
\end{minted}

The first and second argument of type \mintinline{haskell}{Edis} is where the
dictionaries going to stay.

To execute a \mintinline{haskell}{Edis} program, simply apply it to
 \mintinline{haskell}{unEdis} to get an ordinary program of type
 \mintinline{haskell}{Redis}, with type information erased.

@@

Now we have \mintinline{haskell}{Edis}, let's make our own version of
\mintinline{haskell}{ping}\footnotemark

@@

\begin{minted}[xleftmargin=1em]{haskell}
ping :: Edis xs xs (Either Reply Status)
ping = Edis Hedis.ping
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
del :: KnownSymbol s
    => Proxy s
    -> Edis xs (Del xs s) (Either Reply Integer)
del key = Edis $ Hedis.del (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
set :: KnownSymbol s
    => Proxy s
    -> ByteString       -- data to store
    -> Edis xs (Set xs s Strings) (Either Reply Status)
set key val = Edis $ Hedis.set (encodeKey key) val
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
set :: (KnownSymbol s, Serialize x)
    => Proxy s
    -> x       -- can be anything, as long as it's serializable
    -> Edis xs (Set xs s (StringOf x)) (Either Reply Status)
set key val = Edis $ Hedis.set (encodeKey key) (encode val)
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
     => Proxy s -> Edis xs xs (Either Reply Integer)
incr key = Edis $ Hedis.incr (encodeKey key)

incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
     => Proxy s -> Double -> Edis xs xs (Either Reply Double)
incrbyfloat key n = Edis $ Hedis.incrbyfloat (encodeKey key) n
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
        => Proxy s
        -> Edis xs xs (Either Reply Integer)
llen key = Edis $ Hedis.llen (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
llen :: (KnownSymbol s, ListOrNX xs s)
        => Proxy s
        -> Edis xs xs (Either Reply Integer)
llen key = Edis $ Hedis.llen (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
declare :: (KnownSymbol s, Member xs s ~ False)
        => Proxy s
        -> Proxy x  -- type of value
        -> Edis xs (Set xs s x) ()
declare s x = Edis $ return ()

renounce :: (KnownSymbol s, Member xs s ~ True)
        => Proxy s -> Edis xs (Del xs s) ()
renounce s = Edis $ return ()

-- empty precondition
start :: Edis ’[] ’[] ()
start = Edis $ return ()
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
main :: IO ()
main = do
    conn <- connect defaultConnectInfo
    result <- runRedis conn $ unEdis $ start
        `bind` \_ ->  declare
                        (Proxy :: Proxy "A")
                        (Proxy :: Proxy Integer)
        `bind` \_ ->  incr     (Proxy :: Proxy "A")
        `bind` \n ->  case n of
            Left  err -> lpush (Proxy :: Proxy "L") 0
            Right n   -> lpush (Proxy :: Proxy "L") n
        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
    print result
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
get :: (KnownSymbol s, Serialize x
     , Just (StringOf x) ~ Get xs s)
    => Proxy s -> Edis xs xs (Either Reply (Maybe x))
\end{minted}

@@

In Hedis such command could easily be expressed with lists of
 \mintinline{haskell}{ByteString}s. But in Edis, things escalate quickly, as
 the keys and values will have to be expressed with \emph{heterogeneous
 lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
 if not impossible.

>>>>

The database system and the package we are targeting are \emph{Redis} and
 \emph{Hedis} respectively, and our implementation is available as \emph{Popcorn}\footnotemark
 on \emph{Hackage}.

\footnotetext{Popcorn, a temporary codename for the purposes of double-blind reviewing}

@@

\begin{itemize}[noitemsep]
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
 language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and constraints of
 an action in types, with indexed monad\cite{indexedmonad} and other language
 extensions such as closed type-families\cite{closedtypefamilies} and
 constraints kinds\cite{constraintskinds}.
\item Popcorn, a package we built for programmers. This package helps programmers
 to write more reliable Redis programs, and also makes Redis polymorphic by
 automatically converting back and forth from values of arbitrary types and
 boring ByteStrings.
\end{itemize}

@@

We define a new datatype \mintinline{haskell}{Popcorn}, which is basically just
 the context \mintinline{haskell}{Redis} indexed with more information in types.
 We make it an instance of \mintinline{haskell}{IMonad}.

\begin{minted}[xleftmargin=1em]{haskell}
newtype Popcorn p q a = Popcorn { unPopcorn :: Redis a }

instance IMonad Popcorn where
    unit = Popcorn . return
    bind m f = Popcorn (unPopcorn m >>= unPopcorn . f )
\end{minted}

The first and second argument of type \mintinline{haskell}{Popcorn} is where the
dictionaries going to stay.

To execute a \mintinline{haskell}{Popcorn} program, simply apply it to
 \mintinline{haskell}{unPopcorn} to get an ordinary program of type
 \mintinline{haskell}{Redis}, with type information erased.

@@

Now we have \mintinline{haskell}{Popcorn}, let's make our own version of
\mintinline{haskell}{ping}\footnotemark

@@

\begin{minted}[xleftmargin=1em]{haskell}
ping :: Popcorn xs xs (Either Reply Status)
ping = Popcorn Hedis.ping
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
del :: KnownSymbol s
    => Proxy s
    -> Popcorn xs (Del xs s) (Either Reply Integer)
del key = Popcorn $ Hedis.del (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
set :: KnownSymbol s
    => Proxy s
    -> ByteString       -- data to store
    -> Popcorn xs (Set xs s Strings) (Either Reply Status)
set key val = Popcorn $ Hedis.set (encodeKey key) val
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
set :: (KnownSymbol s, Serialize x)
    => Proxy s
    -> x       -- can be anything, as long as it's serializable
    -> Popcorn xs (Set xs s (StringOf x)) (Either Reply Status)
set key val = Popcorn $ Hedis.set (encodeKey key) (encode val)
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
incr :: (KnownSymbol s, Get xs s ~ Just (StringOf Integer))
     => Proxy s -> Popcorn xs xs (Either Reply Integer)
incr key = Popcorn $ Hedis.incr (encodeKey key)

incrbyfloat :: (KnownSymbol s, Get xs s ~ Just (StringOf Double))
     => Proxy s -> Double -> Popcorn xs xs (Either Reply Double)
incrbyfloat key n = Popcorn $ Hedis.incrbyfloat (encodeKey key) n
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
llen :: (KnownSymbol s, Get xs s ~ Just (ListOf x))
        => Proxy s
        -> Popcorn xs xs (Either Reply Integer)
llen key = Popcorn $ Hedis.llen (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
llen :: (KnownSymbol s, ListOrNX xs s)
        => Proxy s
        -> Popcorn xs xs (Either Reply Integer)
llen key = Popcorn $ Hedis.llen (encodeKey key)
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
declare :: (KnownSymbol s, Member xs s ~ False)
        => Proxy s
        -> Proxy x  -- type of value
        -> Popcorn xs (Set xs s x) ()
declare s x = Popcorn $ return ()

renounce :: (KnownSymbol s, Member xs s ~ True)
        => Proxy s -> Popcorn xs (Del xs s) ()
renounce s = Popcorn $ return ()

-- empty precondition
start :: Popcorn ’[] ’[] ()
start = Popcorn $ return ()
\end{minted}

@@

\begin{minted}[xleftmargin=1em]{haskell}
main :: IO ()
main = do
    conn <- connect defaultConnectInfo
    result <- runRedis conn $ unPopcorn $ start
        `bind` \_ ->  declare
                        (Proxy :: Proxy "A")
                        (Proxy :: Proxy Integer)
        `bind` \_ ->  incr     (Proxy :: Proxy "A")
        `bind` \n ->  case n of
            Left  err -> lpush (Proxy :: Proxy "L") 0
            Right n   -> lpush (Proxy :: Proxy "L") n
        `bind` \_ ->  lpop     (Proxy :: Proxy "L")
    print result
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
get :: (KnownSymbol s, Serialize x, StringOrNX xs s)
    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
\end{minted}

@@

\begin{minted}[xleftmargin=1em,fontsize=\footnotesize]{haskell}
get :: (KnownSymbol s, Serialize x
     , Just (StringOf x) ~ Get xs s)
    => Proxy s -> Popcorn xs xs (Either Reply (Maybe x))
\end{minted}

@@

In Hedis such command could easily be expressed with lists of
 \mintinline{haskell}{ByteString}s. But in Popcorn, things escalate quickly, as
 the keys and values will have to be expressed with \emph{heterogeneous
 lists}\cite{hetero}, which would be pratically infeasible, considering the cost,
 if not impossible.
